---
title: "Predicting Risk of Self-Harm and Victimisation Following TBI: R code"
subtitle: "For this project, I utilised Cox regression to investigate the association between paediatric traumatic brain injury (TBI) and subsequent risk of self-harm and violent victimisation, and assess how sociodemographic and clinical factors modify this association. This document outlines the key steps in the analyses and has been prepared in accordance with [SAIL Databank's](https://saildatabank.com) output review policy. <br>

I would like to acknowledge the use of code from Plana-Ripoll and colleagues ([available here](https://plana-ripoll.github.io/NB-COMO/)) that was used guide data cleaning and preparation for the time-to-event analysis."

format:
  lumo-html: 
    #logo:  #optional: add your logo to the doc header
    primary-color: "#b5cef7"
    github-repo: "https://github.com/mayaogonah" 
    self-contained: true
    is-particlejs-enabled: true  # optional: display interactive particles in the doc header
    #bg-image:  # optional: add a background image
    code-fold: show

execute:
  eval: false
  echo: true
author: Maya Ogonah, University of Oxford
date: today
---

# Load Packages

```{r}
# packages installed:
#install.packages("biostat3")
#install.packages("RODBC")
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("lubridate")
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("hrbrthemes")
#install.packages("shiny")
#install.packages("htmlwidgets")
#install.packages("devtools")
#install.packages("car")
#install.packages("pandoc")
#install.packages("modelsummary")
#install.packages("cowplot")
#install.packages(c("systemfonts", "gdtools"))
#install.packages("flextable", type = "binary")
#install.packages("survival")
#install.packages("xfun")
#install.packages("stringr")
#install.packages("survminer")
#install.packages("scales")
#install.packages("geepack")
#install.packages("reshape2")
#install.packages("MetBrewer")
#install.packages ("broom")
#installing the SAILDBUtils Library
devtools::install_git(
  "https://gitlab.sail.ukserp.ac.uk/SAIL/0612/sail-db-utils.git",
  git="external",
  subdir="0612 - SAIL DB Utils/SAILDBUtils"
)

# load packages 
library(biostat3)
library(RODBC)
library(dplyr)
library(tidyr)
library(lubridate)
library(tidyverse)
library(survival)
library(SAILDBUtils)
library(ggplot2)
library(hrbrthemes)
library(shiny)
library(htmlwidgets)
library(devtools)
library(car)
library(modelsummary)
library(cowplot)
library(pandoc)
library(flextable)
library(survival)
library(survminer)
library(stringr)
library(ggplot2)
library(scales)
library(geepack)
library(MetBrewer)
library(broom)
library(reshape2)
library(sf)
library(openxlsx)
pacman::p_load(rms)
library(SAILDBUtils) # this is a package specific to SAIL Databank

# connecting to the SAIL gateway 
# login to DB2 
# test connection by loading the linked file made in SQL
```

# SAIL Databank Geographical Coverage 

```{r}
# load data
gp.data <- read.xlsx("geo-data/gp_coverage.xlsx")

my_geojson <- read_sf("geo-data/features.json")

# plot basic geospatial object
ggplot(my_geojson) + 
  geom_sf(fill = "white", color = "black", linewidth = 0.3) +
  theme_void()

# merge geospatial and numeric data
gp.map <- my_geojson %>%
  left_join(gp.data, by = c("name_en" = "name_en"))

# create the basic choropleth map ###
ggplot(gp.map) + 
  geom_sf(aes(fill = sail_percent)) +
  theme_void()

# customise the plot
plot <- ggplot(gp.map) + 
  geom_sf(aes(fill = sail_percent), linewidth = 0.5, alpha = 0.9) +
  theme_void()+
  scale_fill_viridis_c(option = "mako",
    trans = "log", breaks = c(1, 30, 40, 50, 60, 70, 80, 100),
    name = "SAIL GP (% by No. of Practices)",
    guide = guide_legend(
      keyheight = unit(3, units = "mm"),
      keywidth = unit(10, units = "mm"),
      label.position = "bottom",
      title.position = "top",
      nrow = 1
    )
  ) +
  labs(
    title = "Coverage of General Practices \nin SAIL Databank",
    caption = "Map contains public secotr information lcensed unde the Open Government License v3.0."
  ) +
  theme(
    text = element_text(color = "#22211d"),
    plot.background = element_rect(fill = "#f5f5f2", color = NA),
    panel.background = element_rect(fill = "#f5f5f2", color = NA),
    legend.background = element_rect(fill = "#f5f5f2", color = NA),
    plot.title = element_text(
      size = 40, hjust = 0.01, color = "#4e4d47",
      margin = margin(
        b = -0.1, t = 0.4, l = 2,
        unit = "cm"
      )
    ),
    plot.caption = element_text(
      size = 6,
      color = "#4e4d47",
      margin = margin(
        b = 0.3, r = -99, t = 0.3,
        unit = "cm"
      )
    ),
    legend.position = c(0.32, 0.01)
  )
plot

pdf("geo-data/gp_coverage_plot.pdf", width = 18, height = 13, pointsize = 20)
  plot
dev.off()
```


# Data Wrangling

Load the linked file created using SQL

```{r}
# load the linked file ...

# convert characters to dates
str(cohort)
cohort$FIRST_START_DATE <- ymd(cohort$FIRST_START_DATE)
cohort$LAST_END_DATE <- ymd(cohort$LAST_END_DATE)

data <- cohort %>%
  dplyr::select(-starts_with("PEDW"), -contains("EDDS"), -starts_with("WLGP"), -starts_with("ADDE"), -contains("GP"), -contains("IP"), -CINW_CPR, -CRCS_CPR, PSYCH_EDDS_DT, PSYCH_EDDS_FLAG, NEURO_EDDS_FLAG, NEURO_EDDS_DT)

head(data)

colnames(data) <- c("alf", "start_date", "end_date", "person_time_days", "wob", "dod", "sex", "mtbi_flag", "mtbi_date", "modsevtbi_flag", "modsevtbi_date", "tbi_unspecified_flag", "tbi_unspecified_date", "tbi_life", "tbi_date", "age_at_tbi", "suicidality_life", "suicidality_date", "pretbi_suicidality", "victimisation_life", "victimisation_date", "pretbi_victimisation", "sud_life", "sud_date", "pretbi_sud", "smi_life", "smi_date", "depression_life", "depression_date", "anxiety_life", "anxiety_date", "psychiatric_life", "psychiatric_date", "pretbi_psychiatric", "adhd_life", "adhd_date", "pretbi_adhd", "autism_life", "autism_date", "pretbi_autism", "conduct_disorder_life", "conduct_disorder_date", "pretbi_conduct_disorder", "learning_difficulties_life", "learning_difficulties_date", "pretbi_learning_difficulties", "epilepsy_life", "epilepsy_date", "cerebral_palsy_life", "cerebral_palsy_date", "migraine_life", "migraine_date", "neurological_life", "neurological_date", "pretbi_neurological", "deprivation", "child_maltreatment_flag", "looked_after_life", "looked_after_date", "pretbi_looked_after", "ed_psych_date", "ed_psych_life", "ed_neuro_life", "ed_neuro_date")


data <- data %>%
  mutate(pretbi_neurodevelopmental = ifelse(pretbi_autism == 1 | pretbi_adhd == 1, 1, 0 )) %>%
  mutate(neurodevelopmental_life = ifelse(autism_life == 1 | adhd_life == 1, 1, 0 ))

# individuals with TBI (all mod-sev) after the end of study 
right_cens <- data %>%
  filter(tbi_date > end_date)

data <- data %>%
  mutate(modsevtbi_date = replace(modsevtbi_date, modsevtbi_date > end_date, as.Date(NA)))

data <- data %>%
  mutate(modsevtbi_flag = ifelse(is.na(modsevtbi_date), 0, 1))

data <- data %>%
  mutate(tbi_life = ifelse(mtbi_flag == 0 & modsevtbi_flag == 0 & tbi_unspecified_flag == 0, 0, 1))

data <- data %>%
  mutate(tbi_date = replace(tbi_date, tbi_life == 0, as.Date(NA)))

# some individuals have a mild/moderate-severe/unspecified TBI date however have been missed from the flag 
data <- data %>%
  mutate(tbi_date = case_when(
    !is.na(mtbi_date) ~ mtbi_date, 
    !is.na(modsevtbi_date) ~ modsevtbi_date,
    !is.na(tbi_unspecified_date) ~ tbi_unspecified_date
  )) %>%
  mutate(tbi_life = ifelse(!is.na(tbi_date), 1, 0))

# a few individuals where the later tbi date was chosen as first TBI

# calculate age at TBI injury (and ensure age is numeric)
data$age_at_tbi <- decimal_date(data$tbi_date) - decimal_date(data$wob)
data$age_at_tbi <- as.numeric(data$age_at_tbi)

# some individuals marked as "TBI" even though the event occurred at over 25y
over25_TBI <- data %>%
  filter(age_at_tbi >= 25 & !is.na(tbi_date)) %>%
  dplyr::select("alf", "wob", "age_at_tbi", "tbi_life", "tbi_date", "mtbi_flag", "mtbi_date", "modsevtbi_flag", "modsevtbi_date", "tbi_unspecified_flag", "tbi_unspecified_date")

# individuals with a TBI at over 25 labelled as 'TBI', recode as 'No TBI

# recode TBI_life to 0
data <- data %>%
  mutate(tbi_life = ifelse(tbi_life == 1 & age_at_tbi > 25, 0, tbi_life))

# recode TBI date to NA
data <- data %>%
  mutate(tbi_date = case_when(
    tbi_life == 0 ~ as.Date(NA),
    tbi_life == 1 ~ tbi_date
  ))

# recode age at TBI to NA
data <- data %>%
  mutate(age_at_tbi = ifelse(tbi_life == 1 & !is.na(age_at_tbi), NA_real_, age_at_tbi))

# create a column for diagnosis (TBI / No TBI) and TBI severity
data <- data %>%
  dplyr::mutate(diagnosis = ifelse(tbi_life == 1, 'tbi', 'no tbi'),
                tbi_severity = case_when(
                  (tbi_date == mtbi_date) ~ "probable mild", 
                  (tbi_date == modsevtbi_date) ~ "probable moderate-severe",
                  (tbi_date == tbi_unspecified_date) ~ "probable mild" )
  )

# recoding severity to compare only those with mild to moderate (excluding unspecified)
data <- data %>%
  mutate(tbi_severity_binary = case_when(
    tbi_severity == "probable mild" ~ 0,
    tbi_severity == "probable moderate-severe" ~ 1
  ))

# recode sex, and create variables for variables used for analysis
data <- data %>%
  mutate(across(sex, as.numeric)) %>%
  mutate(sex = replace(sex, sex ==1, 0)) %>%
  mutate(sex = replace(sex, sex ==2, 1))

# flip the coding of deprivation so high deprivation is 'baseline'. A high HR would indicate that high deprivation increases the risk
data$deprivation <-  car::recode(data$deprivation, "5=1; 4=2; 3=3; 2=4; 1=5")
```

## Specifying Follow-up

### Left Truncation

```{r}
# follow-up starts on date of first TBI + 1 (first_tbi_date) 
# for controls follow up starts on entry into the study + 1 (start_date) ###

data <- data %>%
  mutate(entry_date = case_when(
    (tbi_life == 0) ~ start_date + days(1),
    (tbi_life == 1) ~ tbi_date + days(1)
  ))

data$entry_dd <- decimal_date(data$entry_date) # create decimal date for day of entry
data$wob_dd <- decimal_date(data$wob) # create decimal date for week of birth
data$age_at_entry <- data$entry_dd - data$wob_dd # create age at entry (decimal date)
summary(data$entry_date)
```

### Restrict to those Under 25y at Baseline

```{r}
data_25y <- data %>%
  filter(age_at_entry < 25)
nrow(data_25y)

tbi25y <- data %>%
  filter(age_at_entry >= 25 & tbi_life == 1)
```

### Remove those with Missing Data

```{r}
# check missing and incorrectly coded data
columns_to_check <- c("alf", "start_date", "end_date", "wob", "sex", "deprivation")
rows_with_missing_data <- data_25y[apply(is.na(data_25y[, columns_to_check]), 1, any), ]

rows_with_missing_data <- rows_with_missing_data %>%
  dplyr::select("alf", "start_date", "end_date", "wob", "sex", "deprivation", "person_time_days")

## individuals with missing data for ALF_PE, start and end dates (for study period), time-person-days, WOB, and deprivation(WIMD)

data_25y <- anti_join(data_25y, rows_with_missing_data, by = "alf")
# created a cohort with no missing data or miscoded dates 
# for ease of using original code 
data <- data_25y
```

### Right Censoring

```{r}
# follow-up ends on October 16 2023, end of GP registration, or death, whichever occurs first

data$last_date <- as.Date("2023-10-16")

data$exit_date <- pmin(data[, "end_date"], data[, "dod"], data[, "last_date"], na.rm = TRUE)
data$exit_dd <- decimal_date(data$exit) # create decimal date for date of exit
data$age_at_exit <- data$exit_dd - data$wob_dd # create age of exit (decimal date)

#summary(data$exit_date)
#table(data$exit_date)
```

### Eliminate those with "Negative" Follow-Up

```{r}
negativefu <- data %>%
  filter(entry_dd > exit_dd) %>%
  dplyr::select("alf", "start_date", "end_date", "tbi_date", "tbi_life", "victimisation_date", "entry_date", "exit_date", "entry_dd", "exit_dd") 

prop.table(table(negativefu$tbi_life))*100

zerofu <- data %>%
  filter(entry_dd == exit_dd) %>%
  dplyr::select("alf", "start_date", "end_date", "tbi_date", "tbi_life", "victimisation_date", "entry_date", "exit_date", "entry_dd", "exit_dd")

prop.table(table(zerofu$tbi_life))*100

short_fu <- data %>%
  filter(as.numeric(difftime(exit_date, entry_date, units = "days")) < 7) %>%
  dplyr::select("alf", "start_date", "end_date", "tbi_date", "tbi_life", "victimisation_date", "entry_date", "exit_date", "entry_dd", "exit_dd") 

prop.table(table(short_fu$tbi_life))*100

data <- data[data$entry_dd < data$exit_dd, ] # eliminate those with zero follow-up time

data <- data[!is.na(data$alf), ]
```

# Set the Outcome: Self-Harm

## Complete Data Wrangling

### Assign the date of Self-Harm as the Outcome

```{r}
# first lets look at suicidality as the outcome of interest 
outcome <- "suicidality_date"

# follow-up needs to finish on suicidality date, for those with this event
data$exit_date <- pmin(data$exit_date, data$suicidality_date, na.rm = TRUE)
data$exit_dd <- decimal_date(data$exit_date) # create decimal date for date of exit
data$age_at_exit <- data$exit_dd - data$wob_dd # create age of exit 

data$selfharm_dd <- decimal_date(data$suicidality_date) # create decimal date for date of suicidality
```

### Break-Ties

For some people, the start of follow-up and suicidality event might occur on the same day (ties) in time-to-event setting, this individuals are followed to outcome, and these people would not be considered to be 'exposed' this would underestimate all associations, and outcome should be moved 1 day to "break these ties"

```{r}
temp <- data

# create a table with all suicidality dates
# load tables created in SQL with all self-harm events recorded in inpatient, GP, and death records

pedw <- pedw %>%
  select("ALF_PE", "START_DATE", "DIAG_CD_1234")

wlgp <- wlgp %>%
  select("ALF_PE", "EVENT_DT", "EVENT_CD")

adde <- adde %>% 
  select("ALF_PE", "DEATH_DT", "DEATHCAUSE_DIAG_UNDERLYING_CD")

colnames(pedw) <- c("alf", "date", "code")
colnames(wlgp) <- c("alf", "date", "code")
colnames(adde) <- c("alf", "date", "code")

combined_data <- bind_rows(pedw, wlgp, adde)

combined_data <- combined_data %>%
  filter(!is.na(alf))

final_table <- combined_data %>%
  group_by(alf) %>%
  mutate(event_number = row_number()) %>%
  ungroup

final_table <- final_table %>%
  pivot_wider(
    id_cols = alf,
    names_from = event_number, 
    values_from = c(date, code),
    names_glue = "event{event_number}_{.value}"
  ) 

date_cols <- grep("date$", colnames(final_table), value = TRUE)

# identify those with 'ties' 
ties <- temp %>%
  filter(exit_dd - entry_dd == 0.000) %>%
  select(alf, start_date, end_date, tbi_date, suicidality_date, entry_date, exit_date)

filtered_events <- lapply(date_cols, function(date_cols) {
  final_table %>%
    select(alf, all_of(date_cols)) %>%
    rename(event_date = all_of(date_cols)) %>%
    filter(event_date > ties$suicidality_date[match(alf, ties$alf)] &
             event_date <= ties$end_date[match(alf, ties$alf)]) %>%
    mutate(source_column = date_cols)
})

filtered_events <- bind_rows(filtered_events)

first_event_dates <- filtered_events %>%
  dplyr::group_by(alf) %>%
  dplyr::summarize(second_suicidality_date = min(event_date, na.rm = TRUE)) %>%
  dplyr::ungroup()

ties_joined <- ties %>%
  left_join(first_event_dates, by = "alf")

colnames(ties_joined)

ties <- ties %>%
  left_join(ties_joined %>% select(alf, second_suicidality_date), by = "alf")

# add the second suicidality date to the temp table
temp <- temp %>%
  left_join(ties_joined %>% select(alf, second_suicidality_date), by = "alf")

# to break ties exclude outcome event, however, individuals remain at-risk during follow-up
temp <- temp %>%
  mutate(suicidality_date = case_when(
    suicidality_date > entry_date ~ as.Date(suicidality_date),
    suicidality_date <= entry_date ~ NA)
  )

## also for the washout period exclude outcome events within 6 days of study entry / 7 days of the TBI date
short_fu <- temp %>%
  filter(as.numeric(difftime(suicidality_date, entry_date, units = "days")) <= 6) %>%
  dplyr::select(alf, dod, start_date, end_date, tbi_date, tbi_life, suicidality_date)

prop.table(table(short_fu$tbi_life))*100

check <- short_fu %>%
  filter(dod == suicidality_date)

temp <- temp %>%
  mutate(suicidality_date = case_when(
    as.numeric(difftime(suicidality_date, entry_date, units = "days")) > 6 ~ as.Date(suicidality_date),
    as.numeric(difftime(suicidality_date, entry_date, units = "days")) <= 6 ~ NA)
  )

# add the next self-harm event during follow-up
temp <- temp %>%
  mutate(suicidality_date = case_when(
    exit_dd - entry_dd == 0.000 ~ as.Date(second_suicidality_date),
    as.numeric(difftime(suicidality_date, entry_date, units = "days")) <= 6 ~ as.Date(second_suicidality_date),
    TRUE ~ as.Date(suicidality_date)))

# update self-harm flag 
temp <- temp %>%
  mutate(suicidality_life = case_when(
    is.na(suicidality_date) ~ 0,
    !is.na(suicidality_date) ~ 1
    
  ))

temp <- temp %>%
  mutate(selfharm_dd = decimal_date(data$suicidality_date))

# update exit date
temp <- temp %>%
  mutate(exit_date = coalesce(end_date, dod, last_date))
temp <- temp %>%
  mutate(exit_date = coalesce(suicidality_date, exit_date))
temp$exit_dd <- decimal_date(temp$exit_date) # create decimal date for date of exit
temp$age_at_exit <- temp$exit_dd - temp$wob_dd 

# X individual for which event was not censored
# manually alter self-harm date for X participant 
# X individual who was born after study entry
```

### Creating the Fail and Outcome Variables

::: callout-note
For participants with a self-harm event before the start of follow-up, now exit will occur before entry
:::

Creating a "fail" variable for censoring where 0 = censored (they exit on the last date of the study period 16 Oct 2023); 1 = event; 80 = study exit wasn't October 16 2023; 90 = death; 91 = death & diagnosis on the same day.

```{r}

temp$fail <- (90 * (!is.na(temp$dod) & (temp$dod == temp$exit_date))) + 
  (80 * (!is.na(temp$exit_date) & (temp$exit_date != temp$last_date) & is.na(temp$dod) & is.na(temp$suicidality_date))) +
  (01 * (!is.na(temp$suicidality_date) & (temp$suicidality_date <= temp$exit_date)))

table(temp$fail)

# create an "outcome" column for suicidality during follow up, where fail == 1 or fail == 91
temp <- temp %>%
  mutate(outcome = if_else(fail == 1 | fail == 91, 1, 0))

table(temp$outcome)
table(!is.na(temp$suicidality_date), temp$outcome) # checking that everyone with an event date had an 'outcome'

# these people are considered 'censored' as their suicidality date is not equal or before to their exit date ###
```

### Specify Tbi Recurrence (Time Varying Covariate)

```{r}

# create a table with all TBI dates (in secondary records)
# load files from SQL

edds_unspec <- edds_unspec %>%
  dplyr::select("ALF_PE", "ADMIN_ARR_DT")

edds_mild <- edds_mild %>%
  dplyr::select("ALF_PE", "ADMIN_ARR_DT")

edds_mod <- edds_mod %>%
  dplyr::select("ALF_PE", "ADMIN_ARR_DT")

pedw_unspec <- pedw_unspec %>%
  dplyr::select("ALF_PE", "START_DATE")

pedw_mild <- pedw_mild %>%
  dplyr::select("ALF_PE", "START_DATE")

pedw_mod <- pedw_mod %>%
  dplyr::select("ALF_PE", "START_DATE")

colnames(edds_unspec) <- c("alf", "date")
colnames(edds_mild) <- c("alf", "date")
colnames(edds_mod) <- c("alf", "date")
colnames(pedw_unspec) <- c("alf", "date")
colnames(pedw_mild) <- c("alf", "date")
colnames(pedw_mod) <- c("alf", "date")

# link with first TBI already identified (primary and secondary records)
# load files from SQL

tbi <- tbi %>%
  dplyr::select("ALF_PE", "STUDY_DATE")

colnames(tbi) <- c("alf", "date")

combined_data <- bind_rows(edds_unspec, edds_mild, edds_mod, pedw_unspec, pedw_mild, pedw_mod, tbi)

# clean environment
rm(list = c("edds_mild", "edds_mod", "edds_unspec", "pedw_mild", "pedw_mod", "pedw_unspec"))

combined_data <- combined_data %>%
  filter(!is.na(alf))

final_table <- combined_data %>%
  group_by(alf) %>%
  mutate(event_number = row_number()) %>%
  ungroup

final_table <- final_table %>%
  pivot_wider(
    id_cols = alf,
    names_from = event_number, 
    values_from = c(date),
    names_glue = "event{event_number}_{.value}"
  )

final_table_long <- final_table %>%
  pivot_longer(
    cols = matches("date$"),
    names_to = "event_type",
    values_to = "event_date"
  ) %>%
  dplyr::select(-contains("code"))

final_table_long <- final_table_long %>%
  filter(!is.na(event_date))

tbi_recurrent_list <- final_table_long

# ensure only the people identified with a TBI under age 25y are included
tbi_recurrent_list <- tbi_recurrent_list %>% 
  filter(alf %in% tbi$alf)
num_unique_alf <- n_distinct(tbi_recurrent_list$alf)
num_unique_alf 

# ensure only events during the study period (for each person) are included
recurrent_merged <- left_join(tbi_recurrent_list, temp, by = "alf")

# review dates outside of the study period 
tbi_remove <- recurrent_merged %>%
  filter(event_date < start_date | event_date > end_date)

tbi_remove <- tbi_remove %>%
  select("alf", "event_type", "event_date", "start_date", "end_date")

# remove the dates outside the study period
recurrent_merged <- recurrent_merged %>%
  filter(event_date >= start_date | event_date <= end_date)

tbi_recurrent_list <- recurrent_merged %>%
  dplyr::select("alf", "event_type", "event_date", "start_date", "end_date")

tbi_recurrent_list <- recurrent_merged %>%
  dplyr::select("alf", "event_type", "event_date")

# only one tbi per date per individual
duplicates <- tbi_recurrent_list %>%
  group_by(alf, event_date) %>%
  filter(n() > 1) %>%
  ungroup()

tbi_recurrent_list <- tbi_recurrent_list %>%
  arrange(alf, event_date) %>%
  distinct(alf, event_date, .keep_all = TRUE)

# convert event_date to numeric 
## the TBI date should be expressed relative to the start of follow-up
## change_time is time since first TBI
tbi_recurrent_list <- tbi_recurrent_list %>%
  left_join(temp %>% dplyr::select(alf, entry_date), by = "alf") %>%
  mutate(change_time = as.numeric(event_date - entry_date))

tbi_recurrent_list <- tbi_recurrent_list %>%
  mutate(change_time = case_when(
    change_time < 0 ~ 0,
    change_time > 0 ~ change_time
  ))

# arrange and create num_tbi variable
tbi_recurrent_list <- tbi_recurrent_list %>%
  arrange(alf, event_date) %>% # events are ordered by alf and date
  group_by(alf) %>% # group by individual
  mutate(
    num_tbi = row_number()
  ) %>%
  ungroup()

#summary(tbi_recurrent_list$num_tbi)
```

### Prepare for Time-to-Event Analysis

Start of follow-up: date of TBI event + 1 (cases), study entry + 1 (controls).

```{r}
temp2 <- temp %>%
  mutate(s_day = entry_dd - entry_dd) %>%
  mutate(follow_up = exit_dd - entry_dd)

# create s_day to assign as tstart (s_day = 0 so can something to subtract from)

summary(temp2$follow_up)

zero_fu <- temp2 %>%
  filter(follow_up == 0.000) %>%
  dplyr::select("alf", "start_date", "end_date", "tbi_date", "tbi_life", "suicidality_date", "entry_date", "exit_date", "exit_dd", "entry_dd", "s_day", "follow_up") # have effectively broken ties

#table(zero_fu$tbi_life)
#prop.table(table(zero_fu$tbi_life))*100

temp2 <- tmerge(temp2, temp2, id = alf, tstart = s_day, tstop = follow_up)
temp2 <- tmerge(temp2, temp2, id = alf, endpoint = event(tstop, outcome))
# tmerge function help creates start,stop datasets with multiple intervals per participants

# use tmerge to add the time-varying covariate of tbi (recurrent)
temp3 <- tmerge(
  temp2, # the previous `tmerge()` output, an initialised dataset with tstart/stop intervals and event information
  tbi_recurrent_list,  # dataset with time-varying information
  id = alf, # identifier column
  tbi_injuries = tdc(change_time, num_tbi) # time-dependent covariate
)
## if event_date > tstop, the covariate is not added (as it has been censored)

temp3 <- temp3 %>%
  mutate(expo_tbi = ifelse(diagnosis == 'tbi', TRUE, FALSE))

temp3 <- temp3 %>%
  mutate(time = tstop - tstart)

summary(temp3$tbi_injuries)
temp3 <- temp3 %>% 
  mutate(tbi_injuries = if_else(tbi_life == 0 & !is.na(tbi_injuries), NA, tbi_injuries))

# small tweaks to temp2 
temp2 <- temp2 %>%
  mutate(expo_tbi = ifelse(diagnosis == 'tbi', TRUE, FALSE))

temp2 <- temp2 %>%
  mutate(time = tstop - tstart)

## add binary recurrence variable 
tbi_recurr <- tbi_recurrent_list %>%
  group_by(alf) %>%
  summarise(num_tbi = max(num_tbi, na.rm = TRUE)) 

temp2 <- temp2 %>%
  left_join(tbi_recurr, by = "alf")

temp2 <- temp2 %>% 
  mutate(num_tbi = if_else(tbi_life == 0 & !is.na(num_tbi), NA, num_tbi))

temp2 <- temp2 %>% 
  mutate(recurrent_tbi_suicidality = case_when(
    num_tbi == 1 ~ 0,
    num_tbi >= 2 ~ 1
  )) # NA if non-TBI, 0 if 1 TBI, 1 if recurrent TBI

# change num_tbi to the person with one tbi however this is a first tbi and after 25

```

### Save Cleaned Self-Harm Data

```{r}
data_survival <- temp2

data_tdc <- temp3

rm(list =c("data", "temp", "temp2", "ties_joined", "zero_fu")) # remove dfs 
```

## Descriptive Statistics: Self-Harm

Load data and calculate the persons at riskf of self-harm, the number of new cases of self-harm across follow-up and the number of participants exposed to TBI.

```{r}
## persons at risk of self-directed violence during follow-up, overall and by sex
nrow(data_survival) 
table(data_survival$sex) 
table(data_survival$diagnosis) 
table(data_survival$diagnosis, data_survival$sex) 

## new cases of self-directed violence during follow-up, overall and by sex
nrow(data_survival[data_survival$endpoint==1, ]) 
table(data_survival[data_survival$endpoint==1, "sex"]) 
table(data_survival[data_survival$endpoint==1, "diagnosis"]) 
table(data_survival[data_survival$endpoint==1, "diagnosis"], data_survival[data_survival$endpoint==1, "sex"]) 

# total exposure 
table(data_survival$tbi_life)
table(data_survival$tbi_severity) 
table(data_survival$tbi_severity_binary)
table(data_survival$recurrent_tbi_suicidality)
data_survival %>% count(num_tbi) 

data_survival$age_at_selfharm <- data_survival$selfharm_dd - data_survival$wob_dd
str(data_survival$age_at_selfharm)
summary(data_survival$age_at_selfharm)

data_survival$age_at_tbi <- decimal_date(data_survival$tbi_date) - decimal_date(data_survival$wob)
data_survival$age_at_tbi <- as.numeric(data_survival$age_at_tbi)

# age at exposure
mild <- data_survival %>%
  filter(tbi_severity == "probable mild")
summary(mild$age_at_tbi) 

mod_sev <- data_survival %>%
  filter(tbi_severity == "probable moderate-severe")
summary(mod_sev$age_at_tbi) 

tbi <- data_survival %>%
  filter(diagnosis == "tbi")
summary(tbi$age_at_tbi) 

no_tbi <- data_survival %>%
  filter(diagnosis == 'no tbi')

## information on follow-up
person_years_fu <- sum(data_survival$person_time_days) / 365.24

summary(data_survival$time) 
sd(data_survival$time) 

summary(tbi$time) 
sd(tbi$time)

summary(no_tbi$time) 
sd(no_tbi$time) 

sum(data_survival$time) 
sum(tbi$time) 
sum(no_tbi$time)
```

```{r}
# sex descript
table(data_survival$sex)
prop.table(table(data_survival$sex))*100 

# birth year descript
assign_decade <- function(week_date) {
  year <- year(week_date)
  decade_start <- floor((year - 1)/10) *10 + 1
  decade_end <- decade_start + 9
  return(paste(decade_start, "-", decade_end))
}

decade_counts <- data_survival %>%
  mutate(decade = assign_decade(wob)) %>%
  group_by(decade) %>%
  summarise(
    count = n(),
    died = sum(!is.na(dod))) %>%
  mutate(
    percent = (count / sum(count))*100,
    died_percent = (died / count)*100
  ) %>%
  mutate(
    percent = number(percent, accuaracy = 0.1),
    died_percent = number(died_percent, accuaracy = 0.1)
  ) %>%
  arrange(decade)

# age at start and end of follow-up descriptives
summary(tbi$age_at_entry)
summary(no_tbi$age_at_entry)

summary(data_survival$age_at_entry) 
summary(data_survival$age_at_exit) 

# descriptive statistics of covariates: sex, deprivation, adverse childhood experience (lifetime), psychiatric conditions (lifetime), neurological conditions (lifetime), substance misuse (lifetime)

# age at baseline by CDC age categories
data_survival$agebaseline <- cut(data_survival$age_at_entry,
                  breaks = c(-Inf, 4, 9, 14, 19, 25),
                  labels = c("0-4", "5-9", "10-14", "15-19", "20-24"),
                  right = FALSE)

datasummary_crosstab(agebaseline ~ diagnosis, data = data_survival)

# deprivation 
datasummary_crosstab(deprivation ~ diagnosis, data = data_survival)

# other characteristics 
data_descriptive <- data_survival %>%
  mutate(sex = ifelse(sex == 0, 'male', 'female'),
         deprivation = ifelse(deprivation == 5, 'high deprivation', 'moderate-low deprivation'),
         'Lifetime substance misuse' = ifelse(!is.na(sud_date), 'yes', 'no'),
         'Lifetime psychiatric comorbidity' = ifelse(psychiatric_life == 1, 'yes', 'no'),
         'Lifetime neurological comorbidity' = ifelse(neurological_life == 1, 'yes', 'no'),
         'Childhood maltreatment' = ifelse(child_maltreatment_flag == 1, 'yes', 'no'),
         'Self-directed violence' = ifelse(endpoint == 1, 'yes', 'no')) %>%
  select(diagnosis, 
         tbi_severity = tbi_severity,
         sex,
         deprivation,
         'Lifetime substance misuse', 
         'Lifetime psychiatric comorbidity',
         'Lifetime neurological comorbidity',
         'Childhood maltreatment')

data_descriptive <- data_descriptive %>%
  select(diagnosis, tbi_severity, sex, deprivation, 'Lifetime substance misuse', 'Lifetime psychiatric comorbidity', 'Lifetime neurological comorbidity', 'Childhood maltreatment')

datasummary_balance(~diagnosis,
                    data = data_descriptive)

# prior psychiatric and neurological disorders descript
tbi_descriptive <- tbi %>%
  mutate(sex = ifelse(sex == 0, 'male', 'female'),
         deprivation = ifelse(deprivation == 5, 'high deprivation', 'moderate-low deprivation'),
         'substance misuse' = ifelse(pretbi_sud == 1, 'yes', 'no'),
         'psychiatric comorbidity' = ifelse(pretbi_psychiatric == 1, 'yes', 'no'),
         'neurological comorbidity' = ifelse(pretbi_neurological == 1, 'yes', 'no'),
         'self-directed violence' = ifelse(endpoint == 1, 'yes', 'no')) %>%
  select(diagnosis, 
         tbi_severity = tbi_severity,
         sex,
         deprivation,
         'substance misuse', 
         'psychiatric comorbidity',
         'neurological comorbidity',
         'self-directed violence')

tbi_descriptive <- tbi_descriptive %>%
  dplyr::select(diagnosis, tbi_severity, 'substance misuse', 'psychiatric comorbidity', 
                'neurological comorbidity')

datasummary_balance(~diagnosis,
                    data = tbi_descriptive, 
                    title = "Pre-TBI comorbidities descriptives")

datasummary_balance(~tbi_severity,
                    data = tbi_descriptive, 
                    title = "Pre-TBI comorbidities descriptives")

# chi-squared test to check whether the proportion of individuals with lifetime psychiatric conditions differs in those with TBI and those without
chisq.test(data_survival$diagnosis, data_survival$psychiatric_life)

datasummary_balance(~tbi_severity,
                    data = data_descriptive)

# association between TBI (exposure) and suicidal behaviour (outcome)
table(data_survival$tbi_life, data_survival$endpoint) 
prop.table(table(data_survival$tbi_life, data_survival$endpoint))*100 

table(data_survival$tbi_severity, data_survival$endpoint)
prop.table(table(data_survival$tbi_severity, data_survival$endpoint))*100

data_descriptive <- data_survival %>%
  mutate(sex = ifelse(sex == 0, 'male', 'female'),
         'Lifetime substance misuse' = ifelse(!is.na(sud_date), 'yes', 'no'),
         'Lifetime psychiatric comorbidity' = ifelse(psychiatric_life == 1, 'yes', 'no'),
         'Lifetime neurological comorbidity' = ifelse(neurological_life == 1, 'yes', 'no'),
         'Childhood maltreatment' = ifelse(child_maltreatment_flag == 1, 'yes', 'no'),
         'Self-directed violence' = ifelse(endpoint == 1, 'yes', 'no')) %>%
  select(diagnosis, 
         tbi_severity = tbi_severity,
         sex,
         'Lifetime substance misuse', 
         'Lifetime psychiatric comorbidity',
         'Lifetime neurological comorbidity',
         'Childhood maltreatment',
         'Self-directed violence')

data_outcome <- data_descriptive %>%
  select(diagnosis, tbi_severity, 'Self-directed violence')

datasummary_balance(~diagnosis,
                    data = data_outcome, 
                    title = "Self-directed violence Events")

datasummary_balance(~tbi_severity,
                    data = data_outcome, 
                    title = "Self-directed violence Events")


summary(data_survival$age_at_selfharm) 
summary(mild$age_at_selfharm) 
summary(mod_sev$age_at_selfharm) 
summary(tbi$age_at_selfharm) 
summary(no_tbi$age_at_selfharm) 

# number censored during follow-up
censored <- data_survival %>%
  group_by(alf) %>%
  arrange(tstop) %>%
  filter(row_number() ==n())
table(censored$fail,censored$endpoint)
```

## Cumulative Incidence and Kaplan-Meier Curves

```{r}
## cumulative incidence is equal to 1 - survival

fit <- survfit( Surv(time, endpoint) ~ diagnosis, data = data_survival)
summary(fit, times = c(0, 1, 5, 10, 14, 15)) # fit time based on follow-up (and max follow-up)
summary(fit, times = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15))

# incidence is inverse of survival (therefore 1 - survival)
plot <- ggsurvplot(fit, conf.int = TRUE, risk.table = TRUE, cumevents = TRUE, cumcensor = TRUE,
                   #pval = TRUE,
                   #pval.coord = c(0.05, 2.5),
                   fun = function(y) (1-y)*100,
                   xlab = 'Time (in years)',
                   xlim = c(0,12),
                   break.x.by = 3,
                   ylab = 'Cumulative incidence (in %)', 
                   censor = FALSE, 
                   
                   legend.title = "Exposure", 
                   legend.labs = c("No TBI diagnosis", "TBI diagnosis"),
                   
                   smooth = TRUE,
                   
                   tables.height = 0.2,
                   tables.theme = theme_cleantable(),
                   
                   palette = c('cadetblue3', 'lightpink3')
)

km_plot <- plot$plot
risk_table <- plot$table
cum_events <- plot$cumevents
cum_censor <- plot$ncensor.plot

plot2 <- cowplot::plot_grid(
  km_plot,
  risk_table,
  cum_events, 
  cum_censor,
  ncol = 1, 
  align = "v",
  rel_heights = c(0.45, 0.15, 0.15, 0.15)
)

padded_plot <- ggdraw() + 
  draw_plot(plot2, x = 0.05, y = 0.05, width = 0.9, height = 0.9)

padded_plot

plot <- ggsurvplot(fit, conf.int = TRUE, risk.table = FALSE, cumevents = FALSE, cumcensor = FALSE,
                   fun = function(y) (1-y)*100,
                   xlab = 'Time (in years)',
                   xlim = c(0,12),
                   break.x.by = 3,
                   ylab = 'Cumulative incidence (in %)',
                   ylim = c(0,15),
                   break.y.by = 3,
                   censor = FALSE, 
                   
                   legend.title = "Exposure", 
                   legend.labs = c("No TBI diagnosis", "TBI diagnosis"),
                   
                   smooth = TRUE,
                   
                   palette = c('cadetblue3', 'lightpink3'),
) 

plot

# smoothed curves in line with SAIL output procedure 
# first, clean-up the diagnosis variable for the legend
data <- data_survival
data <- data %>%
  mutate(diagnosis = case_when(
    diagnosis == "tbi" ~ "TBI diagnosis",
    diagnosis == "no tbi" ~ "No TBI diagnosis"
  ))
colnames(data)[colnames(data) == "diagnosis"] <- "Exposure"

fit <- survfit( Surv(time, endpoint) ~ Exposure, data = data)

# convert the survival object to a tidy frame
surv_data <- surv_summary(fit, data = data)

# transform survival probabilities to cumulative incidence
surv_data$incidence <- (1-surv_data$surv) * 100 

surv_data <- surv_data %>%
  mutate(strata = case_when(
    strata == "Exposure=No TBI diagnosis" ~ "No TBI diagnosis",
    strata == "Exposure=TBI diagnosis" ~ "TBI diagnosis"
  ))

# plot KM curves with smoothing
plot_smoothed <- ggplot(surv_data, aes(x = time, y = incidence, color = strata))+
  #geom_step(aes(linetype = strata), size = 0.8) + # original KM curve, comment this line out if happy with the smoothing, uncomment to see the steps
  geom_smooth(method = "loess", span = 0.3, se = FALSE, aes(fill = strata), alpha = 0.2)+
  labs(#title = "Cumulative incidence of self-directed violence over time from start of follow-up",
    x = 'Time (in years)',
    y = 'Cumulative incidence (in %)',
    color = "Exposure",
    linetype = "Exposure",
    fill = "Exposure") +
  scale_y_continuous(limits = c(0, 8)) + # scales y axis from 0 to 20 you can adjust as you like. +
  scale_x_continuous(limits = c(0, 14), breaks = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14)) +
  theme_cowplot(12)

plot_smoothed
```

## Cox Regressions

### Model Assumptions

Assess non-linearity of continuous predictors (age) using restricted cubic splines

```{r}
# age at baseline (full sample)
agebaseline3_spline <- rcs(data_survival$age_at_entry,3)
agebaseline4_spline <- rcs(data_survival$age_at_entry,4)
agebaseline5_spline <- rcs(data_survival$age_at_entry,5)

cox_baseline3 <- coxph(Surv(tstart,tstop, endpoint) ~ agebaseline3_spline, data = data_survival, ties = "breslow")
cox_baseline3
lp_baseline3 <- predict (cox_baseline3)

cox_baseline4 <- coxph(Surv(tstart,tstop, endpoint) ~ agebaseline4_spline, data = data_survival, ties = "breslow")
cox_baseline4
lp_baseline4 <- predict (cox_baseline4)

cox_baseline5 <- coxph(Surv(tstart,tstop, endpoint) ~ agebaseline5_spline, data = data_survival, ties = "breslow")
cox_baseline5
lp_baseline5 <- predict (cox_baseline5)

data_splines <- data.frame(data_survival$age_at_entry,lp_baseline3, lp_baseline4, lp_baseline5)
data_splines_m <- melt(data_splines,id.vars = 'data_survival.age_at_entry')
plot_splines <- ggplot(data_splines_m, aes(x=data_survival.age_at_entry, y=value, colour=variable)) + geom_line() + scale_colour_manual(labels = c("3 knots", "4 knots", "5 knots"), values = c('cadetblue3', "darkseagreen", "deeppink3")) + theme_bw()
plot_splines + labs(x = "Age at Baseline", y = "Linear prediction", color = "") + theme(legend.position.inside = c(0.8,0.2))

AIC(cox_baseline3) 
BIC(cox_baseline3) 

AIC(cox_baseline4) 
BIC(cox_baseline4) 

AIC(cox_baseline5) 
BIC(cox_baseline5) 

# age at first TBI
cox_age_tbi <- coxph(Surv(tstart,tstop, endpoint) ~ age_at_tbi, data = tbi)
martingale_residuals <- resid(cox_age_tbi, type = "martingale")

plot(tbi$age_at_tbi, martingale_residuals, pch = 20, col = "cadetblue4",
     xlab = "Age at first TBI", ylab = "Martingale Residual", 
     main = "Residuals vs. TBI Age")

abline(h = 0)
lines(smooth.spline(tbi$cox_age_tbi, martingale_residuals, df = 7), lty = 2, lwd = 2)
dev.off()

ageTBI3_spline <- rcs(tbi$age_at_tbi,3)
ageTBI4_spline <- rcs(tbi$age_at_tbi,4)
ageTBI5_spline <- rcs(tbi$age_at_tbi,5)

cox_ageTBI3 <- coxph(Surv(tstart,tstop, endpoint) ~ ageTBI3_spline, data = tbi, ties = "breslow")
cox_ageTBI3
lp_ageTBI3 <- predict (cox_ageTBI3)

cox_ageTBI4 <- coxph(Surv(tstart,tstop, endpoint) ~ ageTBI4_spline, data = tbi, ties = "breslow")
cox_ageTBI4
lp_ageTBI4 <- predict (cox_ageTBI4)

cox_ageTBI5 <- coxph(Surv(tstart,tstop, endpoint) ~ ageTBI5_spline, data = tbi, ties = "breslow")
cox_ageTBI5
lp_ageTBI5 <- predict (cox_ageTBI5)

data_splines <- data.frame(tbi$age_at_tbi,lp_ageTBI3, lp_ageTBI4, lp_ageTBI5)
data_splines_m <- melt(data_splines,id.vars = 'tbi.age_at_tbi')
plot_splines <- ggplot(data_splines_m, aes(x=tbi.age_at_tbi, y=value, colour=variable)) + geom_line() + scale_colour_manual(labels = c("3 knots", "4 knots", "5 knots"), values = c('cadetblue3', "darkseagreen", "deeppink3")) + theme_bw()
plot_splines + labs(x = "Age at first TBI", y = "Linear prediction", color = "") + theme(legend.position.inside = c(0.8,0.2))

AIC(cox_ageTBI3) 
BIC(cox_ageTBI3) 

AIC(cox_ageTBI4) 
BIC(cox_ageTBI4) 

AIC(cox_ageTBI5) 
BIC(cox_ageTBI5) 

tbi$agetbi <- cut(tbi$age_at_tbi,
                  breaks = c(-Inf, 4, 9, 14, 19, 25),
                  labels = c("0-4", "5-9", "10-14", "15-19", "20-24"),
                  right = FALSE)

```

Assess the proportional hazards assumption. The log-minus-log plot is used to check proportional hazards, if the lines for different groups are approximately parallel, proportional hazards assumption holds. If the lines cross or are not parallel, it suggests that the proportional hazards assumption might be violated.

```{r}
fit <- survfit(Surv(tstart,tstop, endpoint)~sex, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("psychiatric disorder: no", "psychiatric disorder: yes"),
                        legend.title = "",
                        legend = c(0.3,0.95),
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p1 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')

fit <- survfit(Surv(tstart,tstop, endpoint)~deprivation, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', "darkseagreen", "indianred4", "deeppink3", 'lightpink3'),
                        legend.labs = c("WIMD 5: least deprived", "WIMD 4", "WIMD 3", "WIMD 2", "WIMD 1: most deprived"),
                        legend.title = "",
                        legend = c(0.3,0.87), 
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p2 <- smooth_plot + # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) +
  scale_color_manual(values = c('cadetblue3', "darkseagreen", "indianred4", "deeppink3", 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')

fit <- survfit(Surv(tstart,tstop, endpoint)~child_maltreatment_flag, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("child maltreatment: no", "child maltreatment: yes"),
                        legend.title = "",
                        legend = c(0.3,0.95),
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p3 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')


fit <- survfit(Surv(tstart,tstop, endpoint)~psychiatric_life, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("psychiatric disorder: no", "psychiatric disorder: yes"),
                        legend.title = "",
                        legend = c(0.3,0.95),
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p4 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')

fit <- survfit(Surv(tstart,tstop, endpoint)~neurological_life, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("neurological disorder: no", "neurological disorder: yes"),
                        legend.title = "",
                        legend = c(0.3,0.95), 
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p5 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')


fit <- survfit(Surv(tstart,tstop, endpoint)~sud_life, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("sud: no", "sud: yes"),
                        legend.title = "",
                        legend = c(0.3,0.95), 
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p6 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')


fit <- survfit(Surv(tstart,tstop, endpoint)~tbi_severity, data = tbi)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("severity: probable mild", "severity: probable moderate-severe"),
                        legend.title = "",
                        legend = c(0.3,0.95), 
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p7 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')

fit <- survfit(Surv(tstart,tstop, endpoint)~agetbi, data = tbi)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', "darkseagreen", "indianred4", "deeppink3", 'lightpink3'),
                        legend.labs = c("age: 0-4y", "age: 5-9y", "age: 10-14y", "age: 15-19y", "age: 20-24y"),
                        legend.title = "",
                        legend = c(0.3,0.87), 
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p9 <- smooth_plot + # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) +
  scale_color_manual(values = c('cadetblue3', "darkseagreen", "indianred4", "deeppink3", 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')


combined_plot <- cowplot::plot_grid(p1, p2, p3, p4, p5, p6, p7, p9, ncol = 3)

combined_plot  


rm(list =c("p1", "p2", "p3", "p4", "p5", "p6", "p7", "p9", "combined_plot"))

```

### Univariate Models

```{r}
# univariate models with categorical variables 
cox_sex <- coxph(Surv(tstart,tstop, endpoint)~ sex, data = data_survival)
cox_deprivation <- coxph(Surv(tstart,tstop, endpoint)~factor(deprivation), data = data_survival)
cox_cpr <- coxph(Surv(tstart,tstop, endpoint)~child_maltreatment_flag, data = data_survival)
cox_psych <- coxph(Surv(tstart,tstop, endpoint)~psychiatric_life, data = data_survival)
cox_neuro <- coxph(Surv(tstart,tstop, endpoint)~neurological_life, data = data_survival)
cox_sud <- coxph(Surv(tstart,tstop, endpoint)~sud_life, data = data_survival)
                 
  # tbi only cohort 
  cox_sex <- coxph(Surv(tstart,tstop, endpoint)~ sex, data = tbi)
  cox_deprivation <- coxph(Surv(tstart,tstop, endpoint)~factor(deprivation), data = tbi)
  cox_cpr <- coxph(Surv(tstart,tstop, endpoint)~child_maltreatment_flag, data = tbi)
  cox_psych <- coxph(Surv(tstart,tstop, endpoint)~psychiatric_life, data = tbi)
  cox_neuro <- coxph(Surv(tstart,tstop, endpoint)~neurological_life, data = tbi)
  cox_sud <- coxph(Surv(tstart,tstop, endpoint)~sud_life, data = tbi)
  cox_sev <- coxph(Surv(tstart,tstop, endpoint)~tbi_severity_binary, data = tbi)
  cox_recurr <- coxph(Surv(tstart,tstop, endpoint)~tbi_recurrence_victim, data = tbi)
  cox_agetbi <- coxph(Surv(tstart,tstop, endpoint)~factor(agetbi), data = tbi)
  
  # controls only cohort
  cox_sex <- coxph(Surv(tstart,tstop, endpoint)~ sex, data = no_tbi)
  cox_deprivation <- coxph(Surv(tstart,tstop, endpoint)~factor(deprivation), data = no_tbi)
  cox_cpr <- coxph(Surv(tstart,tstop, endpoint)~child_maltreatment_flag, data = no_tbi)
  cox_psych <- coxph(Surv(tstart,tstop, endpoint)~psychiatric_life, data = no_tbi)
  cox_neuro <- coxph(Surv(tstart,tstop, endpoint)~neurological_life, data = no_tbi)
  cox_sud <- coxph(Surv(tstart,tstop, endpoint)~sud_life, data = no_tbi)

```

### Multivariate Models: Total Effect

The total effect of TBI diagnosis on time to self-harm is estimated by Cox proportional hazards regression (survival). Covariates are included to adjust for potential confounders.

```{r}
# model 1: minimially adjusted (sex and birth date)
model_1 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life + sex + rcs(age_at_entry, 5), data = data_survival)
summary(model_1)

print(paste0("HR = ", format(exp(summary(model_1)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_1)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_1)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))

# model 2: model 1 + deprivation 
model_2 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life + sex + rcs(age_at_entry, 5) + factor(deprivation), data = data_survival)
summary(model_2)

print(paste0("HR = ", format(exp(summary(model_2)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_2)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_2)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))

# model 3: model 2 + SUD
model_3 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life + sex + rcs(age_at_entry, 5) + factor(deprivation) + sud_life, data = data_survival)
summary(model_3)

print(paste0("HR = ", format(exp(summary(model_3)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_3)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_3)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))

# model 4: model 3 + childhood maltreatment
model_4 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life + sex + rcs(age_at_entry, 5) + factor(deprivation) + sud_life + child_maltreatment_flag, data = data_survival)
summary(model_4)

print(paste0("HR = ", format(exp(summary(model_4)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_4)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_4)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))


# risk factors (and their HR) in the final multivariable model
result <- summary(model_4)
summary_table <- data.frame(
  Covariate = rownames(result$coefficients),
  HR = result$coefficients[, "exp(coef)"],
  `Lower 95% CI` = result$conf.int[, "lower .95"],
  `Upper 95% CI` = result$conf.int[, "upper .95"],
  `p-value` = result$coefficients[, "Pr(>|z|)"]
)

print(summary_table)
```

### Stratified Analysis

```{r}
# the * operator between tbi_life and psychiatric_life creates an interaction term, meaning the model will estimate separate effects for psychiatric_life == 0 and == 1
# the lincom() command produces a HR for each strata
# therefore, syntac-wise is it an interaction term, but this enables us to run a stratified analysis
table(data_survival$psychiatric_life) 
table(data_survival[data_survival$endpoint==1, "psychiatric_life"]) 

# strata: psychiatric conditions on fully adjusted model
model_5 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life * psychiatric_life + sex + rcs(age_at_entry, 5) + factor(deprivation) + sud_life + child_maltreatment_flag, data = data_survival)

lincom(model_5, c("tbi_life"),eform = TRUE)

lincom(model_5, c("tbi_life + tbi_life:psychiatric_life"),eform = TRUE)

# strata: neurological conditions on fully adjusted model
table(data_survival$neurological_life) 
table(data_survival[data_survival$endpoint==1, "neurological_life"]) 

model_6 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life * neurological_life + sex + rcs(age_at_entry, 5) + factor(deprivation) + sud_life + child_maltreatment_flag, data = data_survival)

lincom(model_6, c("tbi_life"),eform = TRUE)

lincom(model_6, c("tbi_life + tbi_life:neurological_life"),eform = TRUE)
```

### Sensitivity Analysis

```{r}
# model 7: re-specified by severity (probable mild versus probable moderate-severe) on fully-adjusted model
model_7 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_severity + sex + rcs(age_at_entry, 5) + factor(deprivation) + sud_life + child_maltreatment_flag, data = tbi)
summary(model_7)

print(paste0("HR = ", format(exp(summary(model_7)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_7)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_7)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))

# model 8: re-specified by recurrent vs single episode (time-varying covariate)

tbi_tdc <- data_tdc %>%
  filter(diagnosis == "tbi")

model_8 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_injuries + sex + rcs(age_at_entry, 5) + factor(deprivation) + sud_life + child_maltreatment_flag, data = tbi_tdc)
summary(model_8)

print(paste0("HR = ", format(exp(summary(model_8)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_8)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_8)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))

# model 9: (nested) interaction effect of age at injury

data_survival$agetbi <- cut(data_survival$age_at_tbi,
                            breaks = c(-Inf, 4, 9, 14, 19, 25),
                            labels = c("0-4", "5-9", "10-14", "15-19", "20-24"),
                            right = FALSE)


model_9 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life + sex + rcs(age_at_entry, 5) + factor(deprivation) + sud_life + child_maltreatment_flag + # main effects 
                   tbi_life:agetbi, # interaction
                 data = data_survival)
summary(model_9)
exp(coef(model_9))

model9_summary <- summary(model_9)
output_table <- model9_summary$coefficients %>%
  as.data.frame() %>%
  rownames_to_column("Term") %>%
  mutate(
    HR = exp(coef),
    `Lower CI` = exp(coef - 1.96 * `se(coef)`),
    `Upper CI` = exp(coef + 1.96 * `se(coef)`)
  ) %>%
  select(Term, coef, HR, `Lower CI`, `Upper CI`, `Pr(>|z|)`)
print(output_table, digits = 3)
# when e+00 , the exponent is zero and the number is actually as presented

summary_data <- tbi %>%
  group_by(agetbi) %>%
  summarise(suicidality_count = sum(endpoint == 1))

plot_count <- ggplot(summary_data, aes(x = agetbi, y = suicidality_count, fill = agetbi)) + 
  geom_bar(stat = "identity") + 
  labs(
    title = "Number of Suicidal Behaviour Events by Age at TBI",
    x = "Age at TBI",
    y = "Count of Suicidal Behaviour Events", 
    fill = "Age at TBI"
  ) +
  theme_minimal() +
  scale_fill_manual(values = met.brewer("Cassatt1"))

plot_count  
```

# Set the Outcome: Violent Victimisation

Now, lets assess the risk of violent victimisation within our cohort.

## Data Wrangling

### Assign the Date of Victimisation as the Outcome

```{r}
# lets look at violent victimisation as the outcome of interest 
outcome <- "victimisation_date"

# follow-up needs to finish on victimisation date, for those with this event
data$exit_date <- pmin(data$exit_date, data$victimisation_date, na.rm = TRUE)
data$exit_dd <- decimal_date(data$exit_date) # create decimal date for date of exit
data$age_at_exit <- data$exit_dd - data$wob_dd # create age of exit 

data$victim_dd <- decimal_date(data$victimisation_date) # create decimal date for date of victimisation

```

### Break Ties

For some people, the start of follow-up and victimisation event might occur on the same day (ties) in time-to-event setting, this individuals are followed to outcome, and these people would not be considered to be 'exposed' this would underestimate all associations, and outcome should be moved 1 day to "break these ties".

```{r}
temp <- data

# create a table with all victimisation dates
# load files from SQL

pedw <- pedw %>%
  dplyr::select("ALF_PE", "START_DATE", "DIAG_CD_1234")

wlgp <- wlgp %>%
  dplyr::select("ALF_PE", "EVENT_DT", "EVENT_CD")

adde <- adde %>% 
  dplyr::select("ALF_PE", "DEATH_DT", "DEATHCAUSE_DIAG_UNDERLYING_CD")

colnames(pedw) <- c("alf", "date", "code")
colnames(wlgp) <- c("alf", "date", "code")
colnames(adde) <- c("alf", "date", "code")

combined_data <- bind_rows(pedw, wlgp, adde)

combined_data <- combined_data %>%
  filter(!is.na(alf))

final_table <- combined_data %>%
  group_by(alf) %>%
  mutate(event_number = row_number()) %>%
  ungroup

final_table <- final_table %>%
  pivot_wider(
    id_cols = alf,
    names_from = event_number, 
    values_from = c(date, code),
    names_glue = "event{event_number}_{.value}"
  )

final_table_long <- final_table %>%
  pivot_longer(
    cols = matches("date$"),
    names_to = "event_type",
    values_to = "event_date"
  ) %>%
  dplyr::select(-contains("code"))

# identify those with 'ties' 

ties <- temp %>%
  filter(exit_dd - entry_dd == 0.000) %>%
  dplyr::select(alf, start_date, end_date, tbi_date, victimisation_date, entry_date, exit_date)

ties_joined <- ties %>%
  left_join(final_table_long, by = "alf")

ties_joined <- ties_joined %>%
  filter(event_date > victimisation_date & event_date <= end_date)

colnames(ties_joined)

ties_joined <- ties_joined %>%
  dplyr::group_by(alf) %>%
  dplyr::summarize(second_victimisation_date = min(event_date, na.rm = TRUE)) %>%
  dplyr::ungroup()

ties <- ties %>%
  left_join(ties_joined, by = "alf")

# add the second victimisation date to the temp table
temp <- temp %>%
  left_join(ties_joined %>% select(alf, second_victimisation_date), by = "alf")

# to break ties exclude outcome event, however, individuals remain at-risk during follow-up
temp <- temp %>%
  mutate(victimisation_date = case_when(
    victimisation_date > entry_date ~ as.Date(victimisation_date),
    victimisation_date <= entry_date ~ NA)
  )

# also for the washout period exclude outcome events within 6 days of study entry / 7 days of the TBI date
short_fu <- temp %>%
  filter(as.numeric(difftime(victimisation_date, entry_date, units = "days")) <= 6) %>%
  dplyr::select(alf, dod, start_date, end_date, tbi_date, tbi_life, victimisation_date)

prop.table(table(short_fu$tbi_life))*100

check <- short_fu %>%
  filter(dod == victimisation_date) # none of these excluded events had a death recorded on the same day

temp <- temp %>%
  mutate(victimisation_date = case_when(
    as.numeric(difftime(victimisation_date, entry_date, units = "days")) > 6 ~ as.Date(victimisation_date),
    as.numeric(difftime(victimisation_date, entry_date, units = "days")) <= 6 ~ NA)
  )

# add the next victimisation event during follow-up
temp <- temp %>%
  mutate(victimisation_date = case_when(
    exit_dd - entry_dd == 0.000 ~ as.Date(second_victimisation_date),
    as.numeric(difftime(victimisation_date, entry_date, units = "days")) <= 6 ~ as.Date(second_victimisation_date),
    TRUE ~ as.Date(victimisation_date)))

# update victimisation flag 
temp <- temp %>%
  mutate(victimisation_life = case_when(
    is.na(victimisation_date) ~ 0,
    !is.na(victimisation_date) ~ 1
  ))

temp <- temp %>%
  mutate(victim_dd = decimal_date(data$victimisation_date))
head(temp)

# update exit date
temp <- temp %>%
  mutate(exit_date = coalesce(end_date, dod, last_date))
temp <- temp %>%
  mutate(exit_date = coalesce(victimisation_date, exit_date))
temp$exit_dd <- decimal_date(temp$exit_date) # create decimal date for date of exit
temp$age_at_exit <- temp$exit_dd - temp$wob_dd 

rm(list = c("adde", "combined_data", "final_table", "final_table_long", "negativefu", "pedw", "short_fu", "ties", "wlgp", "zerofu"))  

# X individual who was born after study entry
# alter manually, change start up to wob
```

### Creating the Fail and Outcome Variables

Creating a "fail" variable for censoring where 0 = censored (they exit on the last date of the study period 16 Oct 2023); 1 = event; 80 = study exit wasn't October 16 2023; 90 = death; 91 = death & diagnosis on the same day.

```{r}

temp$fail <- (90 * (!is.na(temp$dod) & (temp$dod == temp$exit_date))) + 
  (80 * (!is.na(temp$exit_date) & (temp$exit_date != temp$last_date) & is.na(temp$dod) & is.na(temp$victimisation_date))) +
  (01 * (!is.na(temp$victimisation_date) & (temp$victimisation_date <= temp$exit_date)))

table(temp$fail)

# create an "outcome" column for victimisation during follow up, where fail == 1 or fail == 91
temp <- temp %>%
  mutate(outcome = if_else(fail == 1 | fail == 91, 1, 0))

table(temp$outcome)
table(!is.na(temp$victimisation_date), temp$outcome) # checking that everyone with an event date had an 'outcome'

```

### Specify TBI Recurrence (Time-Varying Covariate)

```{r}
# create a table with all TBI dates (in secondary records)
# load files from SQL

edds_unspec <- edds_unspec %>%
  dplyr::select("ALF_PE", "ADMIN_ARR_DT")
edds_unspec$severity <- "unspecified"

edds_mild <- edds_mild %>%
  dplyr::select("ALF_PE", "ADMIN_ARR_DT")
edds_mild$severity <- "mild"

edds_mod <- edds_mod %>%
  dplyr::select("ALF_PE", "ADMIN_ARR_DT")
edds_mod$severity <- "moderate-severe"

pedw_unspec <- pedw_unspec %>%
  dplyr::select("ALF_PE", "START_DATE")
pedw_unspec$severity <- "unspecified"

pedw_mild <- pedw_mild %>%
  dplyr::select("ALF_PE", "START_DATE")
pedw_mild$severity <- "mild"

pedw_mod <- pedw_mod %>%
  dplyr::select("ALF_PE", "START_DATE")
pedw_mod$severity <- "moderate-severe"

colnames(edds_unspec) <- c("alf", "date", "severity")
colnames(edds_mild) <- c("alf", "date", "severity")
colnames(edds_mod) <- c("alf", "date", "severity")
colnames(pedw_unspec) <- c("alf", "date", "severity")
colnames(pedw_mild) <- c("alf", "date", "severity")
colnames(pedw_mod) <- c("alf", "date", "severity")

# link with first TBI already identified (primary and secondary records)
# load files from SQL

tbi <- tbi %>%
  dplyr::select("ALF_PE", "STUDY_DATE")

colnames(tbi) <- c("alf", "date")

combined_data <- bind_rows(edds_unspec, edds_mild, edds_mod, pedw_unspec, pedw_mild, pedw_mod, tbi)

# clean environment
rm(list = c("edds_mild", "edds_mod", "edds_unspec", "pedw_mild", "pedw_mod", "pedw_unspec"))

combined_data <- combined_data %>%
  filter(!is.na(alf))

final_table <- combined_data %>%
  group_by(alf) %>%
  mutate(event_number = row_number()) %>%
  ungroup

final_table <- final_table %>%
  pivot_wider(
    id_cols = alf,
    names_from = event_number, 
    values_from = c(date),
    names_glue = "event{event_number}_{.value}"
  )

final_table_long <- final_table %>%
  pivot_longer(
    cols = matches("date$"),
    names_to = "event_type",
    values_to = "event_date"
  ) %>%
  dplyr::select(-contains("code"))

final_table_long <- final_table_long %>%
  filter(!is.na(event_date))

tbi_recurrent_list <- final_table_long

# ensure only the people identified with a TBI under age 25y are included
tbi_recurrent_list <- tbi_recurrent_list %>% 
  filter(alf %in% tbi$alf)
num_unique_alf <- n_distinct(tbi_recurrent_list$alf)
num_unique_alf 

# ensure only events during the study period (for each person) are included
recurrent_merged <- left_join(tbi_recurrent_list, temp, by = "alf")

# review dates outside of the study period 
tbi_remove <- recurrent_merged %>%
  filter(event_date < start_date | event_date > end_date)

tbi_remove <- tbi_remove %>%
  select("alf", "event_type", "event_date", "start_date", "end_date")

# remove the dates outside the study period
recurrent_merged <- recurrent_merged %>%
  filter(event_date >= start_date | event_date <= end_date)

tbi_recurrent_list <- recurrent_merged %>%
  dplyr::select("alf", "event_type", "event_date", "start_date", "end_date")

tbi_recurrent_list <- recurrent_merged %>%
  dplyr::select("alf", "event_type", "event_date")

# only one tbi per date per individual
duplicates <- tbi_recurrent_list %>%
  group_by(alf, event_date) %>%
  filter(n() > 1) %>%
  ungroup()

tbi_recurrent_list <- tbi_recurrent_list %>%
  arrange(alf, event_date) %>%
  distinct(alf, event_date, .keep_all = TRUE)

# convert event_date to numeric 
## the TBI date should be expressed relative to the start of follow-up
## change_time is time since first TBI
tbi_recurrent_list <- tbi_recurrent_list %>%
  left_join(temp %>% dplyr::select(alf, entry_date), by = "alf") %>%
  mutate(change_time = as.numeric(event_date - entry_date))

tbi_recurrent_list <- tbi_recurrent_list %>%
  mutate(change_time = case_when(
    change_time < 0 ~ 0,
    change_time > 0 ~ change_time
  ))

# arrange and create num_tbi variable
tbi_recurrent_list <- tbi_recurrent_list %>%
  arrange(alf, event_date) %>% # events are ordered by alf and date
  group_by(alf) %>% # group by individual
  mutate(
    num_tbi = row_number()
  ) %>%
  ungroup()

#summary(tbi_recurrent_list$num_tbi)
```

### Prepare for Time to Event Analysis

Start of follow-up: date of TBI event + 1 (cases), study entry + 1 (controls).

```{r}
# timescale: time since start of follow-up (in years)
# create new variables: id (~alf), tstart (~entry_dd), tstop (~exit_dd) and endpoint (~outcome)

temp2 <- temp %>%
  mutate(s_day = entry_dd - entry_dd) %>%
  mutate(follow_up = exit_dd - entry_dd)

# create s_day to assign as tstart (s_day = 0 so can something to subtract from)

summary(temp2$follow_up)

zero_fu <- temp2 %>%
  filter(follow_up == 0.000) %>%
  dplyr::select("alf", "start_date", "end_date", "tbi_date", "tbi_life", "victimisation_date", "entry_date", "exit_date", "exit_dd", "entry_dd", "s_day", "follow_up") # have effectively broken ties

# tmerge function help creates start,stop datasets with multiple intervals per participants
temp2 <- tmerge(temp2, temp2, id = alf, tstart = s_day, tstop = follow_up)
temp2 <- tmerge(temp2, temp2, id = alf, endpoint = event(tstop, outcome))

# use tmerge to add the time-varying covariate of tbi (recurrent)
temp3 <- tmerge(
  temp2, # the previous `tmerge()` output, an initialised dataset with tstart/stop intervals and event information
  tbi_recurrent_list,  # dataset with time-varying information
  id = alf, # identifier column
  tbi_injuries = tdc(change_time, num_tbi) # time-dependent covariate
) ## if event_date > tstop, the covariate is not added (as it has been censored)

temp3 <- temp3 %>%
  mutate(expo_tbi = ifelse(diagnosis == 'tbi', TRUE, FALSE))

temp3 <- temp3 %>%
  mutate(time = tstop - tstart)

summary(temp3$tbi_injuries)
temp3 <- temp3 %>% 
  mutate(tbi_injuries = if_else(tbi_life == 0 & !is.na(tbi_injuries), NA, tbi_injuries))

# small tweaks to temp2 
temp2 <- temp2 %>%
  mutate(expo_tbi = ifelse(diagnosis == 'tbi', TRUE, FALSE))

temp2 <- temp2 %>%
  mutate(time = tstop - tstart)

## add binary recurrence variable 
tbi_recurr <- tbi_recurrent_list %>%
  group_by(alf) %>%
  summarise(num_tbi = max(num_tbi, na.rm = TRUE)) 

temp2 <- temp2 %>%
  left_join(tbi_recurr, by = "alf")

temp2 <- temp2 %>% 
  mutate(num_tbi = if_else(tbi_life == 0 & !is.na(num_tbi), NA, num_tbi))

temp2 <- temp2 %>% 
  mutate(recurrent_tbi_victimisation = case_when(
    num_tbi == 1 ~ 0,
    num_tbi >= 2 ~ 1
  )) # NA if non-TBI, 0 if 1 TBI, 1 if recurrent TBI

# change num_tbi to the person with one tbi however this is a first tbi and after 25

```

### Save Cleaned Victimisation Data

```{r}
data_survival <- temp2

data_tdc <- temp3

rm(list =c("data", "temp", "temp2", "ties_joined", "zero_fu")) # remove dfs
```

## Descriptive Statistics

Descriptive statistics of violent victimisation as later disorder.

```{r}
## persons at risk of violent victimisation during follow-up, overall and by sex
nrow(data_survival) 
table(data_survival$sex) 
table(data_survival$diagnosis)
table(data_survival$diagnosis, data_survival$sex) 

## new cases of violent victimisation during follow-up, overall and by sex
nrow(data_survival[data_survival$endpoint==1, ]) 
table(data_survival[data_survival$endpoint==1, "sex"]) 
table(data_survival[data_survival$endpoint==1, "diagnosis"]) 
table(data_survival[data_survival$endpoint==1, "diagnosis"], data_survival[data_survival$endpoint==1, "sex"]) 

# total exposure 
table(data_survival$tbi_life) 
table(data_survival$tbi_severity) 
table(data_survival$tbi_severity_binary)
table(data_survival$recurrent_tbi_victimisation)
data_survival %>% count(num_tbi) 

data_survival$age_at_victim <- data_survival$victim_dd - data_survival$wob_dd
str(data_survival$age_at_victim)

data_survival$age_at_tbi <- decimal_date(data_survival$tbi_date) - decimal_date(data_survival$wob)
data_survival$age_at_tbi <- as.numeric(data_survival$age_at_tbi)

# age at exposure
mild <- data_survival %>%
  filter(tbi_severity == "probable mild")
summary(mild$age_at_tbi) 

mod_sev <- data_survival %>%
  filter(tbi_severity == "probable moderate-severe")
summary(mod_sev$age_at_tbi) 

tbi <- data_survival %>%
  filter(diagnosis == "tbi")
summary(tbi$age_at_tbi) 

no_tbi <- data_survival %>%
  filter(diagnosis == 'no tbi')

## information on follow-up
person_years_fu <- sum(data_survival$person_time_days) / 365.24 

summary(data_survival$time) 
sd(data_survival$time) 

summary(tbi$time) 
sd(tbi$time) 

summary(no_tbi$time) 
sd(no_tbi$time) 

sum(data_survival$time) 
sum(tbi$time)
sum(no_tbi$time) 
```

```{r}
# sex descript
table(data_survival$sex)
prop.table(table(data_survival$sex))*100 

# birth year descript
assign_decade <- function(week_date) {
  year <- year(week_date)
  decade_start <- floor((year - 1)/10) *10 + 1
  decade_end <- decade_start + 9
  return(paste(decade_start, "-", decade_end))
}

decade_counts <- data_survival %>%
  mutate(decade = assign_decade(wob)) %>%
  group_by(decade) %>%
  summarise(
    count = n(),
    died = sum(!is.na(dod))) %>%
  mutate(
    percent = (count / sum(count))*100,
    died_percent = (died / count)*100
  ) %>%
  mutate(
    percent = number(percent, accuaracy = 0.1),
    died_percent = number(died_percent, accuaracy = 0.1)
  ) %>%
  arrange(decade)

# age at start and end of follow-up descript
summary(tbi$age_at_entry)
summary(no_tbi$age_at_entry)

summary(tbi$age_at_exit) 
summary(no_tbi$age_at_exit)


# descriptive statistics of covariates: sex, high deprivation, adverse childhood experience (lifetime), psychiatric conditions (lifetime), neurological conditions (lifetime), substance misuse (lifetime)

# sample characteristics
data_survival$agebaseline <- cut(data_survival$age_at_entry,
                  breaks = c(-Inf, 4, 9, 14, 19, 25),
                  labels = c("0-4", "5-9", "10-14", "15-19", "20-24"),
                  right = FALSE)

datasummary_crosstab(agebaseline ~ diagnosis, data = data_survival)

datasummary_crosstab(deprivation ~ diagnosis, data = data_survival)

data_descriptive <- data_survival %>%
  mutate(sex = ifelse(sex == 0, 'male', 'female'),
         'Lifetime substance misuse' = ifelse(!is.na(sud_date), 'yes', 'no'),
         'Lifetime psychiatric condition' = ifelse(psychiatric_life == 1, 'yes', 'no'),
         'Lifetime neurological condition' = ifelse(neurological_life == 1, 'yes', 'no'),
         'Childhood maltreatment' = ifelse(child_maltreatment_flag == 1, 'yes', 'no'),
         'Violent victimisation' = ifelse(endpoint == 1, 'yes', 'no')) %>%
  select(diagnosis, 
         tbi_severity = tbi_severity,
         sex,
         'Lifetime substance misuse', 
         'Lifetime psychiatric condition',
         'Lifetime neurological condition',
         'Childhood maltreatment')

data_descriptive <- data_descriptive %>%
  select(diagnosis, tbi_severity, sex, 'Lifetime substance misuse', 'Lifetime psychiatric condition', 
         'Lifetime neurological condition', 'Childhood maltreatment')

datasummary_balance(~diagnosis,
                    data = data_descriptive)

data_psych <- data_survival %>%
  mutate('ED Psychiatric' = ifelse(ed_psych_life == 1, 'yes', 'no'),
         'SMI' =ifelse(smi_life == 1, 'yes', 'no'),
         'Anxiety' =ifelse(anxiety_life == 1, 'yes', 'no'),
         'Depression' =ifelse(depression_life == 1, 'yes', 'no'),
         'ADHD' = ifelse(adhd_life == 1, 'yes', 'no'),
         'Conduct disorder' = ifelse(conduct_disorder_life == 1, 'yes', 'no')) %>%
  select(diagnosis, 
         tbi_severity, 
         'ED Psychiatric', 
         'SMI',
         'Anxiety',
         'Depression', 
         'ADHD', 
         'Conduct disorder')

datasummary_balance(~diagnosis,
                    data = data_psych)

data_neuro <- data_survival %>%
  mutate('ED Neurological' = ifelse(ed_neuro_life == 1, 'yes', 'no'),
         'Epilepsy' =ifelse(epilepsy_life == 1, 'yes', 'no'),
         'Migraine' =ifelse(migraine_life == 1, 'yes', 'no'),
         'Cerebral palsy' = ifelse(cerebral_palsy_life == 1, 'yes', 'no')) %>%
  select(diagnosis, 
         tbi_severity,
         'ED Neurological',
         'Epilepsy',
         'Migraine',
         'Cerebral palsy')

datasummary_balance(~diagnosis,
                    data = data_neuro)

datasummary_crosstab(agebaseline ~ tbi_severity, data = data_survival)

datasummary_crosstab(deprivation ~ tbi_severity, data = data_survival)

datasummary_balance(~tbi_severity,
                    data = data_descriptive) 

datasummary_balance(~tbi_severity,
                    data = data_neuro)

# prior psychiatric and neurological disorders descript
tbi_descriptive <- tbi %>%
  mutate(sex = ifelse(sex == 0, 'male', 'female'),
         deprivation = ifelse(deprivation == 5, 'high deprivation', 'moderate-low deprivation'),
         'substance misuse' = ifelse(pretbi_sud == 1, 'yes', 'no'),
         'psychiatric comorbidity' = ifelse(pretbi_psychiatric == 1, 'yes', 'no'),
         'neurological comorbidity' = ifelse(pretbi_neurological == 1, 'yes', 'no'),
         'violent victimisation' = ifelse(endpoint == 1, 'yes', 'no')) %>%
  select(diagnosis, 
         tbi_severity = tbi_severity,
         sex,
         deprivation,
         'substance misuse', 
         'psychiatric comorbidity',
         'neurological comorbidity',
         'violent victimisation')

tbi_descriptive <- tbi_descriptive %>%
  select(diagnosis, tbi_severity, 'substance misuse', 'psychiatric comorbidity', 
         'neurological comorbidity')

datasummary_balance(~diagnosis,
                    data = tbi_descriptive, 
                    title = "Pre-TBI comorbidities descriptives")

datasummary_balance(~tbi_severity,
                    data = tbi_descriptive, 
                    title = "Pre-TBI comorbidities descriptives")

# association between TBI (exposure) and violent victimisation (outcome)
table(data_survival$tbi_life, data_survival$endpoint)
prop.table(table(data_survival$tbi_life, data_survival$endpoint))*100 

table(data_survival$tbi_severity, data_survival$endpoint) # of those exposed to probable mild TBI 2261 were victimised, 209 of those probable mod-severe TBI were victimised
prop.table(table(data_survival$tbi_severity, data_survival$endpoint))*100

data_descriptive <- data_survival %>%
  mutate(sex = ifelse(sex == 0, 'male', 'female'),
         deprivation = ifelse(deprivation == 5, 'high deprivation', 'moderate-low deprivation'),
         'Lifetime substance misuse' = ifelse(!is.na(sud_date), 'yes', 'no'),
         'Lifetime psychiatric comorbidity' = ifelse(psychiatric_life == 1, 'yes', 'no'),
         'Lifetime neurological comorbidity' = ifelse(neurological_life == 1, 'yes', 'no'),
         'Childhood maltreatment' = ifelse(child_maltreatment_flag == 1, 'yes', 'no'),
         'violent victimisation' = ifelse(endpoint == 1, 'yes', 'no')) %>%
  select(diagnosis, 
         tbi_severity = tbi_severity,
         sex,
         deprivation,
         'Lifetime substance misuse', 
         'Lifetime psychiatric comorbidity',
         'Lifetime neurological comorbidity',
         'Childhood maltreatment',
         'violent victimisation')

data_outcome <- data_descriptive %>%
  select(diagnosis, tbi_severity, 'violent victimisation')

datasummary_balance(~diagnosis,
                    data = data_outcome, 
                    title = "Violent Victimisation Events")

datasummary_balance(~tbi_severity,
                    data = data_outcome, 
                    title = "Violent Victimisation Events")


summary(data_survival$age_at_victim) 
summary(mild$age_at_victim) 
summary(mod_sev$age_at_victim) 
summary(tbi$age_at_victim) 
summary(no_tbi$age_at_victim) 

# number censored during follow-up
censored <- data_survival %>%
  group_by(alf) %>%
  arrange(tstop) %>%
  filter(row_number() ==n())
table(censored$fail,censored$endpoint)

```

## Cumulative Incidence and Kaplan-Meier Curves

```{r}
# cumulative incidence is equal to 1 - survival

fit <- survfit( Surv(time, endpoint) ~ diagnosis, data = data_survival)
summary(fit, times = c(0, 1, 5, 10, 14, 15)) # fit time based on follow-up (and max follow-up)
summary(fit, times = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15))

# incidence is inverse of survival (therefore 1 - survival)
plot <- ggsurvplot(fit, conf.int = TRUE, risk.table = TRUE, cumevents = TRUE, cumcensor = TRUE,
                   #pval = TRUE,
                   #pval.coord = c(0.05, 2.5),
                   fun = function(y) (1-y)*100,
                   xlab = 'Time (in years)',
                   xlim = c(0,12),
                   break.x.by = 3,
                   ylab = 'Cumulative incidence (in %)', 
                   censor = FALSE, 
                   
                   legend.title = "Exposure", 
                   legend.labs = c("No TBI diagnosis", "TBI diagnosis"),
                   
                   smooth = TRUE,
                   
                   tables.height = 0.2,
                   tables.theme = theme_cleantable(),
                   
                   palette = c('cadetblue3', 'lightpink3')
)

km_plot <- plot$plot
risk_table <- plot$table
cum_events <- plot$cumevents
cum_censor <- plot$ncensor.plot

plot2 <- cowplot::plot_grid(
  km_plot,
  risk_table,
  cum_events, 
  cum_censor,
  ncol = 1, 
  align = "v",
  rel_heights = c(0.45, 0.15, 0.15, 0.15)
)

padded_plot <- ggdraw() + 
  draw_plot(plot2, x = 0.05, y = 0.05, width = 0.9, height = 0.9)

padded_plot

plot <- ggsurvplot(fit, conf.int = TRUE, risk.table = FALSE, cumevents = FALSE, cumcensor = FALSE,
                   fun = function(y) (1-y)*100,
                   xlab = 'Time (in years)',
                   xlim = c(0,12),
                   break.x.by = 3,
                   ylab = 'Cumulative incidence (in %)',
                   ylim = c(0,15),
                   break.y.by = 3,
                   censor = FALSE, 
                   
                   legend.title = "Exposure", 
                   legend.labs = c("No TBI diagnosis", "TBI diagnosis"),
                   
                   smooth = TRUE,
                   
                   palette = c('cadetblue3', 'lightpink3'),
) 

plot


# create a smoothed KM curve that aligns with SAIL output review policy
data <- data_survival
data <- data %>%
  mutate(diagnosis = case_when(
    diagnosis == "tbi" ~ "TBI diagnosis",
    diagnosis == "no tbi" ~ "No TBI diagnosis"
  ))
colnames(data)[colnames(data) == "diagnosis"] <- "Exposure"

fit <- survfit( Surv(time, endpoint) ~ Exposure, data = data)

# convert the survival object to a tidy frame
surv_data <- surv_summary(fit, data = data)

# transform survival probabilities to cumulative incidence
surv_data$incidence <- (1-surv_data$surv) * 100 

surv_data <- surv_data %>%
  mutate(strata = case_when(
    strata == "Exposure=No TBI diagnosis" ~ "No TBI diagnosis",
    strata == "Exposure=TBI diagnosis" ~ "TBI diagnosis"
  ))

# plot KM curves with smoothing
plot_smoothed <- ggplot(surv_data, aes(x = time, y = incidence, color = strata))+
  #geom_step(aes(linetype = strata), size = 0.8) + # original KM curve, comment this line out if happy with the smoothing, uncomment to see the steps
  geom_smooth(method = "loess", span = 0.3, se = FALSE, aes(fill = strata), alpha = 0.2)+
  labs(#title = "Cumulative incidence of violent victimisation over time from start of follow-up",
    x = 'Time (in years)',
    y = 'Cumulative incidence (in %)',
    color = "Exposure",
    linetype = "Exposure",
    fill = "Exposure") +
  scale_y_continuous(limits = c(0, 4.5)) + # scales y axis from 0 to 20 you can adjust as you like. +
  scale_x_continuous(limits = c(0, 14), breaks = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14)) +
  theme_cowplot(12)

plot_smoothed
```

## Cox Regressions

### Model Assumptions

Assess non-linearity of continuous predictors.

```{r}
# age at baseline
agebaseline3_spline <- rcs(data_survival$age_at_entry,3)
agebaseline4_spline <- rcs(data_survival$age_at_entry,4)
agebaseline5_spline <- rcs(data_survival$age_at_entry,5)

cox_baseline3 <- coxph(Surv(tstart,tstop, endpoint) ~ agebaseline3_spline, data = data_survival, ties = "breslow")
cox_baseline3
lp_baseline3 <- predict (cox_baseline3)

cox_baseline4 <- coxph(Surv(tstart,tstop, endpoint) ~ agebaseline4_spline, data = data_survival, ties = "breslow")
cox_baseline4
lp_baseline4 <- predict (cox_baseline4)

cox_baseline5 <- coxph(Surv(tstart,tstop, endpoint) ~ agebaseline5_spline, data = data_survival, ties = "breslow")
cox_baseline5
lp_baseline5 <- predict (cox_baseline5)

data_splines <- data.frame(data_survival$age_at_entry,lp_baseline3, lp_baseline4, lp_baseline5)
data_splines_m <- melt(data_splines,id.vars = 'data_survival.age_at_entry')
plot_splines <- ggplot(data_splines_m, aes(x=data_survival.age_at_entry, y=value, colour=variable)) + geom_line() + scale_colour_manual(labels = c("3 knots", "4 knots", "5 knots"), values = c('cadetblue3', "darkseagreen", "deeppink3")) + theme_bw()
plot_splines + labs(x = "Age at Baseline", y = "Linear prediction", color = "") + theme(legend.position.inside = c(0.8,0.2))


AIC(cox_baseline3) 
BIC(cox_baseline3) 

AIC(cox_baseline4) 
BIC(cox_baseline4) 

AIC(cox_baseline5) 
BIC(cox_baseline5) 

# age at first TBI
cox_age_tbi <- coxph(Surv(tstart,tstop, endpoint) ~ age_at_tbi, data = tbi)
martingale_residuals <- resid(cox_age_tbi, type = "martingale")

plot(tbi$age_at_tbi, martingale_residuals, pch = 20, col = "cadetblue4",
     xlab = "Age at first TBI", ylab = "Martingale Residual", 
     main = "Residuals vs. TBI Age")

abline(h = 0)
lines(smooth.spline(tbi$cox_age_tbi, martingale_residuals, df = 7), lty = 2, lwd = 2)
dev.off()

ageTBI3_spline <- rcs(tbi$age_at_tbi,3)
ageTBI4_spline <- rcs(tbi$age_at_tbi,4)
ageTBI5_spline <- rcs(tbi$age_at_tbi,5)

cox_ageTBI3 <- coxph(Surv(tstart,tstop, endpoint) ~ ageTBI3_spline, data = tbi, ties = "breslow")
cox_ageTBI3
lp_ageTBI3 <- predict (cox_ageTBI3)

cox_ageTBI4 <- coxph(Surv(tstart,tstop, endpoint) ~ ageTBI4_spline, data = tbi, ties = "breslow")
cox_ageTBI4
lp_ageTBI4 <- predict (cox_ageTBI4)

cox_ageTBI5 <- coxph(Surv(tstart,tstop, endpoint) ~ ageTBI5_spline, data = tbi, ties = "breslow")
cox_ageTBI5
lp_ageTBI5 <- predict (cox_ageTBI5)

pdf("S:/1478 - Risk of violence perpetration and victimisation among individuals with traum/output/epi/victimisation/plots/rcs/ageTBI-splines-new.pdf")
data_splines <- data.frame(tbi$age_at_tbi,lp_ageTBI3, lp_ageTBI4, lp_ageTBI5)
data_splines_m <- melt(data_splines,id.vars = 'tbi.age_at_tbi')
plot_splines <- ggplot(data_splines_m, aes(x=tbi.age_at_tbi, y=value, colour=variable)) + geom_line() + scale_colour_manual(labels = c("3 knots", "4 knots", "5 knots"), values = c('cadetblue3', "darkseagreen", "deeppink3")) + theme_bw()
plot_splines + labs(x = "Age at first TBI", y = "Linear prediction", color = "") + theme(legend.position.inside = c(0.8,0.2))
dev.off()

AIC(cox_ageTBI3) 
BIC(cox_ageTBI3) 

AIC(cox_ageTBI4) 
BIC(cox_ageTBI4) 

AIC(cox_ageTBI5) 
BIC(cox_ageTBI5) 

tbi$agetbi <- cut(tbi$age_at_tbi,
                  breaks = c(-Inf, 4, 9, 14, 19, 25),
                  labels = c("0-4", "5-9", "10-14", "15-19", "20-24"),
                  right = FALSE)

```

Proportional hazards assumption

```{r}
# check: is there a pattern with time? a systematic departure from a horizontal line is indicative of non-proportional hazards
## multi-level factors will account for banding
## appears to be proportional hazards

fit <- survfit(Surv(tstart,tstop, endpoint)~sex, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("psychiatric disorder: no", "psychiatric disorder: yes"),
                        legend.title = "",
                        legend = c(0.3,0.95),
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p1 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')

fit <- survfit(Surv(tstart,tstop, endpoint)~deprivation, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', "darkseagreen", "indianred4", "deeppink3", 'lightpink3'),
                        legend.labs = c("WIMD 5: least deprived", "WIMD 4", "WIMD 3", "WIMD 2", "WIMD 1: most deprived"),
                        legend.title = "",
                        legend = c(0.3,0.87), 
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p2 <- smooth_plot + # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) +
  scale_color_manual(values = c('cadetblue3', "darkseagreen", "indianred4", "deeppink3", 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')

fit <- survfit(Surv(tstart,tstop, endpoint)~child_maltreatment_flag, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("child maltreatment: no", "child maltreatment: yes"),
                        legend.title = "",
                        legend = c(0.3,0.95),
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p3 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')


fit <- survfit(Surv(tstart,tstop, endpoint)~psychiatric_life, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("psychiatric disorder: no", "psychiatric disorder: yes"),
                        legend.title = "",
                        legend = c(0.3,0.95),
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p4 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')

fit <- survfit(Surv(tstart,tstop, endpoint)~neurological_life, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("neurological disorder: no", "neurological disorder: yes"),
                        legend.title = "",
                        legend = c(0.3,0.95), 
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p5 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')


fit <- survfit(Surv(tstart,tstop, endpoint)~sud_life, data = data_survival)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("sud: no", "sud: yes"),
                        legend.title = "",
                        legend = c(0.3,0.95), 
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p6 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')


fit <- survfit(Surv(tstart,tstop, endpoint)~tbi_severity, data = tbi)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', 'lightpink3'), 
                        legend.labs = c("severity: probable mild", "severity: probable moderate-severe"),
                        legend.title = "",
                        legend = c(0.3,0.95), 
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p7 <- smooth_plot +  # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) + 
  scale_color_manual(values = c('cadetblue3', 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')


fit <- survfit(Surv(tstart,tstop, endpoint)~agetbi, data = tbi)
surv_plot <- ggsurvplot(fit,
                        fun = "cloglog",
                        palette = c('cadetblue3', "darkseagreen", "indianred4", "deeppink3", 'lightpink3'),
                        legend.labs = c("age: 0-4y", "age: 5-9y", "age: 10-14y", "age: 15-19y", "age: 20-24y"),
                        legend.title = "",
                        legend = c(0.3,0.87), 
                        censor = FALSE
)

smooth_plot <- surv_plot$plot
smooth_plot$layers <- smooth_plot$layers[-1]

p9 <- smooth_plot + # smooth the plot
  geom_smooth(method = "loess", 
              span = 0.3, 
              se = FALSE, 
              alpha = 0.2, 
              aes(color = strata),
              linetype = "solid",
              linewidth = 2, ) +
  scale_color_manual(values = c('cadetblue3', "darkseagreen", "indianred4", "deeppink3", 'lightpink3')) + 
  labs(
    x = 'Time',
    y = 'Log(-log(Survival Probability))')


combined_plot <- cowplot::plot_grid(p1, p2, p3, p4, p5, p6, p7, p9, ncol = 3)

combined_plot  


rm(list =c("p1", "p2", "p3", "p4", "p5", "p6", "p7", "p8", "p9", "combined_plot"))

## the log-minus-log plot is used to check proportional hazards, if the lines for different groups are approximately parallel, proportional hazards assumption holds. If the lines cross or are not parallel, it suggests that the proportional hazards assumption might be violated

```

### Univariate Models

```{r}
cox_sex <- coxph(Surv(tstart,tstop, endpoint)~ sex, data = data_survival)
cox_deprivation <- coxph(Surv(tstart,tstop, endpoint)~factor(deprivation), data = data_survival)
cox_cpr <- coxph(Surv(tstart,tstop, endpoint)~child_maltreatment_flag, data = data_survival)
cox_psych <- coxph(Surv(tstart,tstop, endpoint)~psychiatric_life, data = data_survival)
cox_neuro <- coxph(Surv(tstart,tstop, endpoint)~neurological_life, data = data_survival)
cox_sud <- coxph(Surv(tstart,tstop, endpoint)~sud_life, data = data_survival)

  # tbi only cohort 
  cox_sex <- coxph(Surv(tstart,tstop, endpoint)~ sex, data = tbi)
  cox_deprivation <- coxph(Surv(tstart,tstop, endpoint)~factor(deprivation), data = tbi)
  cox_cpr <- coxph(Surv(tstart,tstop, endpoint)~child_maltreatment_flag, data = tbi)
  cox_psych <- coxph(Surv(tstart,tstop, endpoint)~psychiatric_life, data = tbi)
  cox_neuro <- coxph(Surv(tstart,tstop, endpoint)~neurological_life, data = tbi)
  cox_sud <- coxph(Surv(tstart,tstop, endpoint)~sud_life, data = tbi)
  cox_sev <- coxph(Surv(tstart,tstop, endpoint)~tbi_severity_binary, data = tbi)
  cox_recurr <- coxph(Surv(tstart,tstop, endpoint)~tbi_recurrence_victim, data = tbi)
  cox_agetbi <- coxph(Surv(tstart,tstop, endpoint)~factor(agetbi), data = tbi)

  # controls only cohort 
  cox_sex <- coxph(Surv(tstart,tstop, endpoint)~ sex, data = no_tbi)
  cox_deprivation <- coxph(Surv(tstart,tstop, endpoint)~factor(deprivation), data = no_tbi)
  cox_cpr <- coxph(Surv(tstart,tstop, endpoint)~child_maltreatment_flag, data = no_tbi)
  cox_psych <- coxph(Surv(tstart,tstop, endpoint)~psychiatric_life, data = no_tbi)
  cox_neuro <- coxph(Surv(tstart,tstop, endpoint)~neurological_life, data = no_tbi)
  cox_sud <- coxph(Surv(tstart,tstop, endpoint)~sud_life, data = no_tbi)

```

### Multivariate Models

```{r}
# model 1: minimially adjusted (sex and birth date)
model_1 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life + sex + rcs(age_at_entry, 4), data = data_survival)
summary(model_1)
# wald statistic (z coloumn) evaluates whether the beta coefficient of a given variable is statistically significant from 0 | diagnosis, sex and wob have stat sig coefficients
# regression coefficients (coef), with a positive sign signifying that the hazard is higher | victimisation is higher for those diagnosed with TBI and for women, hazard risk decreases with age (wob)
# global statistical significance of the model (likelihood ratio test, wald test, logrank test) | sig model 

print(paste0("HR = ", format(exp(summary(model_1)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_1)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_1)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))

# model 2: model 1 + deprivation 
model_2 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life + sex + rcs(age_at_entry, 4) + factor(deprivation), data = data_survival)
summary(model_2)

print(paste0("HR = ", format(exp(summary(model_2)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_2)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_2)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))


# model 3: model 2 + SUD
model_3 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life + sex + rcs(age_at_entry, 4) + factor(deprivation) + sud_life, data = data_survival)
summary(model_3)

print(paste0("HR = ", format(exp(summary(model_3)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_3)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_3)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))

# model 4: model 3 + childhood maltreatment
model_4 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life + sex + rcs(age_at_entry, 4) + factor(deprivation) + sud_life + child_maltreatment_flag, data = data_survival)
summary(model_4)

print(paste0("HR = ", format(exp(summary(model_4)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_4)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_4)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))

# risk factors (and their HR) in the final multivariable model
result <- summary(model_4)
summary_table <- data.frame(
  Covariate = rownames(result$coefficients),
  HR = result$coefficients[, "exp(coef)"],
  `Lower 95% CI` = result$conf.int[, "lower .95"],
  `Upper 95% CI` = result$conf.int[, "upper .95"],
  `p-value` = result$coefficients[, "Pr(>|z|)"]
)

print(summary_table)
```

### Stratified Analyses

```{r}
# the * operator between tbi_life and psychiatric_life creates an interaction term, meaning the model will estimate separate effects for psychiatric_life == 0 and == 1
# the lincom() command produces a HR for each strata
# therefore, syntac-wise is it an interaction term, but this enables us to run a stratified analysis
table(data_survival$psychiatric_life)
table(data_survival[data_survival$endpoint==1, "psychiatric_life"])

# strata: psychiatric conditions on fully adjusted model
model_5 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life * psychiatric_life + sex + rcs(age_at_entry, 4) + factor(deprivation) + sud_life + child_maltreatment_flag, data = data_survival)

lincom(model_5, c("tbi_life"),eform = TRUE)

lincom(model_5, c("tbi_life + tbi_life:psychiatric_life"),eform = TRUE)

# strata: neurological conditions on fully adjusted model
table(data_survival$neurological_life) 
table(data_survival[data_survival$endpoint==1, "neurological_life"]) 

model_6 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life * neurological_life + sex + rcs(age_at_entry, 4) + factor(deprivation) + sud_life + child_maltreatment_flag, data = data_survival)

lincom(model_6, c("tbi_life"),eform = TRUE)

lincom(model_6, c("tbi_life + tbi_life:neurological_life"),eform = TRUE)

```

### Sensitivity Analysis

```{r}
# model 7: re-specified by severity (probable mild versus probable moderate-severe) on fully-adjusted model
model_7 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_severity + sex + rcs(age_at_entry, 5) + factor(deprivation) + sud_life + child_maltreatment_flag, data = tbi)
summary(model_7)

print(paste0("HR = ", format(exp(summary(model_7)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_7)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_7)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))

# model 8: re-specified by recurrent vs single episode
tbi_tdc <- data_tdc %>%
  filter(tbi_life == 1)

model_8 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_injuries + sex + rcs(age_at_entry, 5) + factor(deprivation) + sud_life + child_maltreatment_flag, data = tbi_tdc)
summary(model_8)

print(paste0("HR = ", format(exp(summary(model_8)$coefficients[1, 1]), digits = 1, nsmall = 1),
             " [95% CI: ", format(summary(model_8)$conf.int[1, 3], digits = 1, nsmall = 1), "-",
             format(summary(model_8)$conf.int[1, 4], digits = 1, nsmall = 1), "]"))
# [1] "HR = 1.6 [95% CI: 1.3-2.0]"

## model 9: (nested) interaction effect of age at injury

data_survival$agetbi <- cut(data_survival$age_at_tbi,
                            breaks = c(-Inf, 4, 9, 14, 19, 25),
                            labels = c("0-4", "5-9", "10-14", "15-19", "20-24"),
                            right = FALSE)

model_9 <- coxph(Surv(tstart,tstop, endpoint) ~ tbi_life + sex + rcs(age_at_entry, 4) + factor(deprivation) + sud_life + child_maltreatment_flag + # main effects 
                   tbi_life:agetbi, # interaction
                 data = data_survival)
summary(model_9)
exp(coef(model_9))

model9_summary <- summary(model_9)
output_table <- model9_summary$coefficients %>%
  as.data.frame() %>%
  rownames_to_column("Term") %>%
  mutate(
    HR = exp(coef),
    `Lower CI` = exp(coef - 1.96 * `se(coef)`),
    `Upper CI` = exp(coef + 1.96 * `se(coef)`)
  ) %>%
  select(Term, coef, HR, `Lower CI`, `Upper CI`, `Pr(>|z|)`)
print(output_table, digits = 3)
# when e+00 , the exponent is zero and the number is actually as presented
# increased risk of violent victimisation as the age at which the injury is sustained increases

plot_scatter <- ggplot(data_survival, aes(x = age_at_tbi, y = age_at_victim)) +
  geom_point(color = "indianred4", size = 3) + 
  geom_smooth(method = "lm", color = 'lightpink3', se = FALSE) + 
  labs(title = "Age at TBI vs. Age at Victimisation",
       x = "Age at TBI", 
       y = "Age at Victimisation") +
  theme_minimal()

plot_scatter


summary_data <- tbi %>%
  group_by(agetbi) %>%
  summarise(victimisation_count = sum(endpoint == 1))

plot_count <- ggplot(summary_data, aes(x = agetbi, y = victimisation_count, fill = agetbi)) + 
  geom_bar(stat = "identity") + 
  labs(
    title = "Number of Victimisation Events by Age at TBI",
    x = "Age at TBI",
    y = "Count of Victimisation Events", 
    fill = "Age at TBI"
  ) +
  theme_minimal() +
  scale_fill_manual(values = met.brewer("Cassatt1"))

plot_count  
```

# Clean Up R Workspace

```{r}
# free unused R memory
gc()

# clear environment ###
rm(list = ls())

# clear console ###
cat("\014")
```
